Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2025-07-10 POPL2026/paper/sections/introduction.tex
one more bib fix

All these issues can be resolved by the use of \emph{string diagrams} in place of textual syntax.
\citet{ghica2024equivalencehypergraphsegraphsmonoidal}~shows how e-graphs (without binding) can be given a categorical semantics using symmetric monoidal semilattice-enriched categories, which allows e-graphs to be reconstructed as string diagrams.
In this paper, we extend this technique to e-graphs with bindings, lifting to monoidal \emph{closed} categories.

@@

The general background literature for string diagrams and their relationship to formal languages is summarised by~\citet{piedeleu2023introductionstringdiagramscomputer}.
The particular context of string diagrams with bindings is given in~\citet{ghica2024stringdiagramslambdacalculifunctional}, while string diagrams for e-graphs without bindings are defined in~\citet{ghica2024equivalencehypergraphsegraphsmonoidal}.

@@

\begin{figure*}[h!]
	\[
		\adjustbox{scale=0.6}{
			\tikzfig{figures/e-graph-example-2}
		}
	\]
	\caption{E-graph example (top) and its equivalent string diagram representation (bottom)~\cite{ghica2024equivalencehypergraphsegraphsmonoidal}}
	\label{fig:e-graph-example}
\end{figure*}

Each column represents a term (or term rewrite rule), its conventional e-graph, and the equivalent string diagram representation using the \emph{e-hypergraphs} of~\citet{ghica2024equivalencehypergraphsegraphsmonoidal}.
For the initial term, $(a*2)/2$, which is already represented efficiently in the e-graph by sharing the node $2$, the string diagram version shares this node explicitly, using a copy node.

@@

\citet{ghica2024equivalencehypergraphsegraphsmonoidal} shows that e-graphs over signature $\Sigma$ can be represented as morphisms of a free cartesian symmetric monoidal semilattice-enriched category over the same signature.
We argue that restricting the categorical domain to free \textbf{closed} symmetric monoidal semilattice-enriched categories gives rise to structures that naturally support binding and equivalence classes of morphisms.
That is, morphisms (string diagrams) of such category can both encode variable binding (using the closed structure) and equivalence (using the semilattice enrichment) between subterms (sub-diagrams).
We call these string diagrams, or more precisely, the combinatorial representation of them, \emph{closed e-hypergraphs}.
First, we define closed enriched $\Sigma$-terms which generate a theory with equivalences and bindings.
Then we define what e-hypergraphs are and provide the interpretation of closed enriched $\Sigma$-terms as cospans of e-hypergraphs (and by doing this we also formalise the corresponding string diagrams as concrete combinatorial objects).
Next we formulate the DPO rewriting for e-hypergraphs and show that this notion of rewriting is sound and complete with respect to rewriting of closed enriched $\Sigma$-terms modulo SMC equations.
We finish the paper with a description of what benefits the e-hypergraph representation may bring for equality saturation in the context of $\lambda$-calculus with explicit substitution compared to conventional (slotted) e-graphs.

>>>>

All these issues can be resolved by the use of \emph{string diagrams} in place of textual syntax.
\citet{tiurin2025equivalencehypergraphsdporewriting}~shows how e-graphs (without binding) can be given a categorical semantics using symmetric monoidal semilattice-enriched categories, which allows e-graphs to be reconstructed as string diagrams.
In this paper, we extend this technique to e-graphs with bindings, lifting to monoidal \emph{closed} categories.

@@

The general background literature for string diagrams and their relationship to formal languages is summarised by~\citet{piedeleu2023introductionstringdiagramscomputer}.
The particular context of string diagrams with bindings is given in~\citet{ghica2024stringdiagramslambdacalculifunctional}, while string diagrams for e-graphs without bindings are defined in~\citet{tiurin2025equivalencehypergraphsdporewriting}.

@@

\begin{figure*}[h!]
	\[
		\adjustbox{scale=0.6}{
			\tikzfig{figures/e-graph-example-2}
		}
	\]
	\caption{E-graph example (top) and its equivalent string diagram representation (bottom)~\cite{tiurin2025equivalencehypergraphsdporewriting}}
	\label{fig:e-graph-example}
\end{figure*}

Each column represents a term (or term rewrite rule), its conventional e-graph, and the equivalent string diagram representation using the \emph{e-hypergraphs} of~\citet{tiurin2025equivalencehypergraphsdporewriting}.
For the initial term, $(a*2)/2$, which is already represented efficiently in the e-graph by sharing the node $2$, the string diagram version shares this node explicitly, using a copy node.

@@

\citet{tiurin2025equivalencehypergraphsdporewriting} shows that e-graphs over signature $\Sigma$ can be represented as morphisms of a free cartesian symmetric monoidal semilattice-enriched category over the same signature.
We argue that restricting the categorical domain to free \textbf{closed} symmetric monoidal semilattice-enriched categories gives rise to structures that naturally support binding and equivalence classes of morphisms.
That is, morphisms (string diagrams) of such category can both encode variable binding (using the closed structure) and equivalence (using the semilattice enrichment) between subterms (sub-diagrams).
We call these string diagrams, or more precisely, the combinatorial representation of them, \emph{closed e-hypergraphs}.
First, we define closed enriched $\Sigma$-terms which generate a theory with equivalences and bindings.
Then we define what e-hypergraphs are and provide the interpretation of closed enriched $\Sigma$-terms as cospans of e-hypergraphs (and by doing this we also formalise the corresponding string diagrams as concrete combinatorial objects).
Next we formulate the DPO rewriting for e-hypergraphs and show that this notion of rewriting is sound and complete with respect to rewriting of closed enriched $\Sigma$-terms modulo SMC equations.
We finish the paper with a description of what benefits the e-hypergraph representation may bring for equality saturation in the context of $\lambda$-calculus with explicit substitution compared to conventional (slotted) e-graphs.

### 2025-07-10 closed_monoidal/closed_monoidal.tex
one more bib fix

We can extend the definition of e-hypergraphs for $\catname{SLat}$-enriched symmetric monoidal categories~\cite{ghica2024equivalencehypergraphsegraphsmonoidal} to accommodate new structure brought by closedness.

>>>>

We can extend the definition of e-hypergraphs for $\catname{SLat}$-enriched symmetric monoidal categories~\cite{tiurin2025equivalencehypergraphsdporewriting} to accommodate new structure brought by closedness.
